---
title: "Adding geological timescales to phylogenies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding geological timescales to phylogenies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, fig.width = 7, fig.height = 5, fig.align = "center")
```

<div style="text-align: justify">
`coord_geo()` is also designed to work smoothly with phylogenies that are plotted with the [ggtree](https://www.bioconductor.org/packages/ggtree) package. First let's load some important packages.

```{r message = FALSE}
# Load deeptime
library(deeptime)
# Load tidyverse (includes ggplot2)
library(tidyverse)
# Load ggtree
library(ggtree)
# Load phytools for some example data
library(phytools)
data(mammal.tree)
# Load paleotree for some example fossil data
library(paleotree)
data(RaiaCopesRule)
```

## Timescales and phylogenies
Now, let's plot some phylogenies with timescales! Note that by default the time axis increases toward the tips for trees plotted with `ggtree()`, so you must use the `revts()` function from [ggtree](https://www.bioconductor.org/packages/ggtree) to change the x-axis values to increasing negative values. Note that once you've done this, you will need to set `neg = TRUE` in `coord_geo()`. Furthermore, any time axis limits that you set will need to be negative. Here we also use the `scale_x_continuous()` function to relabel the negative x-axis values with positive labels.
```{r}
p <- ggtree(mammal.tree) +
  coord_geo(xlim = c(-75, 0), ylim = c(-2, Ntip(mammal.tree)), neg = TRUE, abbrv = FALSE) +
  scale_x_continuous(breaks=seq(-80, 0, 20), labels=abs(seq(-80, 0, 20))) +
  theme_tree2()
revts(p)
```

### Phylogenies with only fossil taxa
Many phylogenies only have fossil taxa in them (i.e., non-ultrametric). These can be handled by `coord_geo()` by using `position_nudge()` function. In most cases, you will want to nudge the time axis values (in this case, the x-axis) by the `root.time` of your phylogeny. Also, note that we have modified `plot.margin` here to accommodate the stacked timescale.
```{r}
ggtree(ceratopsianTreeRaia, position = position_nudge(x = -ceratopsianTreeRaia$root.time)) +
  coord_geo(xlim = c(-163.5, -66), ylim = c(-2, Ntip(ceratopsianTreeRaia)),
            pos = list("bottom", "bottom"), skip = c("Paleocene", "Middle Jurassic"),
            dat = list("epochs", "periods"), abbrv = FALSE,
            size = list(4, 5), neg = TRUE, center_end_labels = TRUE) +
  scale_x_continuous(breaks = -rev(epochs$max_age), labels = rev(epochs$max_age)) +
  theme_tree2() +
  theme(plot.margin = margin(7, 11, 7, 11))
```

## Circular phylogenies
Phylogenies can be plotted in a circular layout with a timescale background using the `coord_geo_polar()` function. It works just like `coord_geo()` and has many of the same arguments. Note that labels for the intervals are not currently implemented (because of the potential mess this could cause if it was automated).
```{r}
revts(ggtree(mammal.tree)) +
  coord_geo_polar(dat = "stages") +
  scale_x_continuous(breaks = seq(-60, 0, 20), labels = abs(seq(-60, 0, 20)))
```

### Circular phylogenies with "stacked" timescales
Timescales can even be "stacked" like with `coord_geo()`, although the stacking occurs in a circular fashion. In this case, we use `scale_y_continuous()` to add a small expansion to the y-axis (which is now the theta axis) so the numbers don't overlap with the phylogeny. We also use the `prop` and `start` arguments to specify how the different timescales are oriented and split within the polar space. The `axis.line.r`, `axis.text.r`, `axis.ticks.r`, and `axis.ticks.length.r` ggplot2 `theme()` elements can be modified just like their x and y counterparts to change the appearance of the radial axis. The default settings work well for a horizontal axis pointing towards the right, but these theme settings will need to be modified for other orientations such as in this example.
```{r message = FALSE}
revts(ggtree(mammal.tree)) +
  coord_geo_polar(dat = list("stages", "periods"), alpha = .5,
                  prop = list(0.66, .34), start = pi/4, lty = "dashed") +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.02))) +
  scale_x_continuous(breaks = seq(-60, 0, 20), labels = abs(seq(-60, 0, 20))) +
  theme(axis.text.r = element_text(size = 3.5, hjust = .25, vjust = .5))
```

### Disclaimer
Note that `coord_geo_polar()` has been tested with the plotting of ultrametric and non-ultrametric phylogenies with `ggtree::ggtree()`, but it may have unexpected behavior when combined with other functions from [ggtree](https://www.bioconductor.org/packages/ggtree). Furthermore, `coord_geo_polar()` may be useful (albeit perhaps abstract) for plotting other types of data, but this has not been investigated.
</div>
