---
title: "Plotting trait data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting trait data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, fig.width = 7, fig.height = 5, fig.align = "center")
```

<div style="text-align: justify">
Many packages exist to visualize trait data for biological species. __deeptime__ similarly has a few novel ways to help you plot your data in useful ways. We'll first load some packages and example data so we can demonstrate some of this functionality.

```{r message = FALSE}
# Load deeptime
library(deeptime)
# Load tidyverse (includes ggplot2)
library(tidyverse)
# Load dispRity for example data
library(dispRity)
data(demo_data)
# Load paleotree for example data
library(phytools)
data(mammal.tree)
data(mammal.data)
```

## Plot disparity through time
A common way to visualize trait data, especially for fossil species, is to show the two-dimensional trait distribution for several time intervals. This allows the viewer to easily compare the trait distribution through time. However, producing such a plot has historically been very time intensive, often involving the use of custom code and image editing software (e.g., [Inkscape](https://inkscape.org/)). While a single function to accomplish such a visualization still does not exist for `{ggplot2}` (yet...), the `coord_trans_xy()` function can be used to generate a similar plot with sheared trait space across several time intervals.

```{r}
#make transformer
library(ggforce)
trans <- linear_trans(shear(.75, 0))

# prepare data to be plotted
crinoids <- as.data.frame(demo_data$wright$matrix[[1]][, 1:2])
crinoids$time <- "before extinction"
crinoids$time[demo_data$wright$subsets$after$elements] <- "after extinction"

# a box to outline the trait space
square <- data.frame(V1 = c(-.6, -.6, .6, .6), V2 = c(-.4, .4, .4, -.4))

ggplot() +
  geom_segment(data = data.frame(x = -.6, y = seq(-.4, .4,.2),
                                 xend = .6, yend = seq(-0.4, .4, .2)),
               aes(x = x, y = y, xend = xend, yend=yend),
               linetype = "dashed", color = "grey") +
  geom_segment(data = data.frame(x = seq(-.6, .6, .2), y = -.4,
                                 xend = seq(-.6, .6, .2), yend = .4),
               aes(x = x, y = y, xend = xend, yend=yend),
               linetype = "dashed", color = "grey") +
  geom_polygon(data = square, aes(x = V1, y = V2), fill = NA, color = "black") +
  geom_point(data = crinoids, aes(x = V1, y = V2), color = 'black') +
  coord_trans_xy(trans = trans, expand = FALSE) +
  labs(x = "PCO1", y = "PCO2") +
  theme_classic() +
  facet_wrap(~time, ncol = 1, strip.position = "right") +
  theme(panel.spacing = unit(1, "lines"), panel.background = element_blank())
```

### Disparity in base R
The `disparity_through_time()` function accomplishes nearly all of the work for you if you are comfortable plotting within the `{lattice}` framework (base R). Note that it may take some tweaking (especially the `aspect` argument) to get the results to look the way you want.

```{r, fig.height = 4}
crinoids$time <- factor(crinoids$time)
disparity_through_time(time~V2*V1, data = crinoids, groups = time, aspect = c(1.5, .6),
                       xlim = c(-.6,.6), ylim = c(-.5,.5),
                       col.regions = "lightyellow", col.point = c("red","blue"),
                       par.settings = list(axis.line = list(col = "transparent"),
                                           layout.heights =
                                               list(top.padding = -20, main.key.padding = 0,
                                                    key.axis.padding = 0, axis.xlab.padding = 0,
                                                    xlab.key.padding = 0, key.sub.padding = 0,
                                                    bottom.padding = -20),
                                           layout.widths =
                                               list(left.padding = -10, key.ylab.padding = 0,
                                                    ylab.axis.padding = 0, axis.key.padding = 0,
                                                    right.padding = 0)))
```

## Phylomorphospaces
Often, trait data will be accompanied with a phylogeny. You may want to visualize both your phylogeny, the traits of your species, and the evolution of the trait along your phylogeny. To accomplish this, you can create a two-dimensional phylomorphospace. The `{phytools}` package has the `phytools::phylomorphospace()` function for accomplishing this in base R. The `geom_phylomorpho()` function in __deeptime__ will help you accomplish this with `ggplot()`. Note that labels can be added using `geom_label()` or `ggrepel::geom_label_repel()`, but they are not demonstrated here because they would obscure the phylogenetic relationships.
```{r message = FALSE, warning = FALSE}
mammal.data <- mammal.data %>%
  rownames_to_column("label")

ggplot(mammal.data, aes(x = bodyMass, y = homeRange, label = label)) +
  geom_phylomorpho(mammal.tree) +
  theme_classic()
```

</div>
